---
title: "Query gene features"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
---
# Loading Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning=FALSE, message = FALSE, dpi = 300
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# install.packages("tidyverse")
# install.packages("BiocManager") #(only if you do not have BiocManager)
#BiocManager::install("org.Hs.eg.db", "biomaRt")

library(tidyverse)
library(biomaRt)
select = dplyr::select
rename = dplyr::rename

set.seed(42)
```

# Human
## Get all gene features
```{r}
# browseVignettes("biomaRt")  # biomart tutorial

# Use biomart to get gene symbols and chromosome start and end positions
mart <- biomaRt::useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                         dataset = "hsapiens_gene_ensembl")
#  mirror = c("uswest", "useast", "asia", "www")

biomaRt::listAttributes(mart)
# biomaRt::listAttributes(mart) %>% filter(str_detect(name, "symbol"))
# biomaRt::listEnsembl()
# biomaRt::searchDatasets(mart = mart, pattern = "hsapiens")
# biomaRt::listFilters(mart)

gene_features <- getBM(
  attributes=c("ensembl_gene_id", "ensembl_gene_id_version", "hgnc_symbol","entrezgene_id","chromosome_name","start_position","end_position"),
  mart = mart)

head(gene_features)
gene_features %>% nrow()
gene_features %>% select(hgnc_symbol) %>% n_distinct()
```
## Filter genes for inferCNV
```{r}
gene_features$chromosome_name %>% unique()

# Filter for genes that have hgnc symbols and are in "main" chromosomes
gene_features.f <- gene_features %>% select(hgnc_symbol, chromosome_name, start_position, end_position) %>% filter(chromosome_name %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,"X", "Y", "MT")) %>% filter(hgnc_symbol != "") %>% unique()

gene_features.f$chromosome_name %>% unique()

nrow(gene_features.f)
gene_features.f$hgnc_symbol %>% n_distinct()
```
InferCNV does not accept duplicated gene features. We are arbitrarily choosing one entry for genes with multiple entries that have (1) the first start position, then (2) the first end position

## Choose entries for duplicate symbols
```{r}
# Genes with duplication
duplicated_genes <- gene_features.f$hgnc_symbol[duplicated(gene_features.f$hgnc_symbol)]
duplicated_genes

# Determine entries with the first start and end positions
gene_features.f <- gene_features.f %>% group_by(hgnc_symbol) %>% mutate(first_start = start_position == min(start_position), first_end = end_position == min(end_position))

# Choose entries with the first start position
duplicated <- gene_features.f %>% filter(hgnc_symbol %in% duplicated_genes)
duplicated <- duplicated %>% filter(first_start)

# Genes that still have duplication
duplicated_genes <- duplicated$hgnc_symbol[duplicated(duplicated$hgnc_symbol)]
duplicated_genes

# From the remaining duplications, choose entries with the first end position
duplicated <- duplicated %>% filter((hgnc_symbol %in% duplicated_genes & first_end) | !hgnc_symbol %in% duplicated_genes)

# There should be no duplicated entries left
duplicated_genes <- duplicated$hgnc_symbol[duplicated(duplicated$hgnc_symbol)]
duplicated_genes

# Choose genes without duplicates, then add the chosen entries
gene_features.f <- gene_features.f %>% 
  filter(!hgnc_symbol %in% duplicated$hgnc_symbol) %>% 
  rbind(duplicated) %>% select(-c(first_start, first_end))

colnames(gene_features.f) <- c("gene_symbol","chromosome_name","start_position","end_position")


nrow(gene_features.f)
```

## Export
```{r}
gene_features %>% write_csv2("biomaRt_gene_features.csv")
gene_features.f %>% write_csv2("inferCNV_gene_features.csv")

rm(duplicated, gene_features, gene_features.f, mart)
```

# Mouse
## Get all gene features
```{r}
# Use biomart to get gene symbols and chromosome start and end positions
mart <- biomaRt::useMart(biomart = "ENSEMBL_MART_ENSEMBL",
                         dataset = "mmusculus_gene_ensembl")
#  mirror = c("uswest", "useast", "asia", "www")

biomaRt::listAttributes(mart)
biomaRt::listAttributes(mart) %>% filter(str_detect(name, "symbol"))
# biomaRt::listEnsembl()
# biomaRt::searchDatasets(mart = mart, pattern = "hsapiens")
# biomaRt::listFilters(mart)

gene_features <- getBM(
  attributes=c("ensembl_gene_id", "ensembl_gene_id_version", "hgnc_symbol", "mgi_symbol", "entrezgene_id","chromosome_name","start_position","end_position"),
  mart = mart)

head(gene_features)
gene_features %>% nrow()
gene_features %>% select(hgnc_symbol) %>% n_distinct()
```
## Filter genes for inferCNV
```{r}
gene_features$chromosome_name %>% unique()

# Filter for genes that have hgnc symbols and are in "main" chromosomes
gene_features.f <- gene_features %>% select(mgi_symbol, chromosome_name, start_position, end_position) %>% filter(chromosome_name %in% c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,"X", "Y", "MT")) %>% filter(mgi_symbol != "") %>% unique()

gene_features.f$chromosome_name %>% unique()

nrow(gene_features.f)
gene_features.f$mgi_symbol %>% n_distinct()
```
InferCNV does not accept duplicated gene features. We are arbitrarily choosing one entry for genes with multiple entries that have (1) the first start position, then (2) the first end position

## Choose entries for duplicate symbols
```{r}
# Genes with duplication
duplicated_genes <- gene_features.f$mgi_symbol[duplicated(gene_features.f$mgi_symbol)]
duplicated_genes

# Determine entries with the first start and end positions
gene_features.f <- gene_features.f %>% group_by(mgi_symbol) %>% mutate(first_start = start_position == min(start_position), first_end = end_position == min(end_position))

# Choose entries with the first start position
duplicated <- gene_features.f %>% filter(mgi_symbol %in% duplicated_genes)
duplicated <- duplicated %>% filter(first_start)

# Genes that still have duplication
duplicated_genes <- duplicated$mgi_symbol[duplicated(duplicated$mgi_symbol)]
duplicated_genes

# From the remaining duplications, choose entries with the first end position
duplicated <- duplicated %>% filter((mgi_symbol %in% duplicated_genes & first_end) | !mgi_symbol %in% duplicated_genes)

# There should be no duplicated entries left
duplicated_genes <- duplicated$mgi_symbol[duplicated(duplicated$mgi_symbol)]
duplicated_genes

# Choose genes without duplicates, then add the chosen entries
gene_features.f <- gene_features.f %>% 
  filter(!mgi_symbol %in% duplicated$mgi_symbol) %>% 
  rbind(duplicated) %>% select(-c(first_start, first_end))

colnames(gene_features.f) <- c("gene_symbol","chromosome_name","start_position","end_position")

nrow(gene_features.f)

```

## Export
```{r}
gene_features %>% write_csv2("biomaRt_gene_features_mouse.csv")
gene_features.f %>% write_csv2("inferCNV_gene_features_mouse.csv")

```

# sessionInfo - Package Versions
```{r}
sessionInfo()
```

