---
title: "Run Pseudobulk CIN DESEq on scRNA-seq dataset"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
  folder: ""
  export: TRUE
  gene_sets: "msigdb_hallmarks_set.csv" 
  
---
# Loading Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning=FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# install.packages("BiocManager") #(only if you do not have BiocManager)
#BiocManager::install("fgsea", "GO.db", "GOstats", "org.Hs.eg.db", "glmGamPoi", "DESeq2")

# install.packages("Seurat", "patchwork", "ggrepel", "sparseMatrixStats", "msigdbr", "tidyverse", "proxy", "plotly", "pheatmap", "gplots")

library(Seurat)
library(patchwork)
library(ggrepel)
library(sparseMatrixStats)
library(msigdbr)
library(org.Hs.eg.db)
library(tidyverse)
library(proxy)
library(fgsea)
library(plotly)
library(pheatmap)
library(gplots)
library(DESeq2)

source("Helper_functions.R")

set.seed(42)
select = dplyr::select
rename = dplyr::rename
```

# Setup
```{r Setup}
folder = params$folder
export = params$export

# generate output dir path named data
source.output.dir <- file.path(file.path(folder, "output_DE"))

# if source output dir does not exist, create it
if (!dir.exists(source.output.dir)) {
    dir.create(source.output.dir)
} else{
  print("Output folder already exists")
}

```

# Load data
```{r Load data}
# Load data
Seurat_object <- readRDS(file = file.path(folder, "output_sc", "Seurat_object_n.rds"))

group_assignment <- read_csv2(file.path(folder, "output_cnv", "CNV_metrics_CIN_assignment.csv"))

Seurat_object[[]]$Project %>% unique()
```

# CIN metadata
```{r CIN metadata}
sc_CIN <- Seurat_object@meta.data %>% select(annotation) %>% left_join(group_assignment %>% rename("annotation" = group), by = "annotation")

# sc_CIN %>% mutate_if(is.character, as.factor) %>% summary()
Seurat_object@meta.data$CIN_status <- sc_CIN$CIN_status
Seurat_object@meta.data$heterogeneity_score <- sc_CIN$heterogeneity_score
Seurat_object@meta.data$aneuploidy_score <- sc_CIN$aneuploidy_score
Seurat_object@meta.data$annotation <-  sc_CIN$annotation
```

```{r}
# Get means of sc expression per gene from "data" and "count" slots
mat <- Seurat::GetAssayData(Seurat_object, assay = "RNA", slot = "data")

sc_means <- mat %>% rowMeans() %>% as.data.frame() %>% rename("sc_mean" = ".") %>% rownames_to_column("symbol")

sc_mean_counts <- Seurat::GetAssayData(Seurat_object, assay = "RNA", slot = "counts") %>% rowMeans() %>% as.data.frame() %>% rename("sc_mean_count" = ".") %>% rownames_to_column("symbol")

sc_means <- sc_means %>% left_join(sc_mean_counts, by = "symbol") %>% mutate(gene = symbol)

if(export){
sc_means %>% write_csv2(file = file.path(source.output.dir, "sc_means.csv"))
}

rm(mat, sc_mean_counts)
```

# Pseudobulk
```{r Pseudobulk}
Seurat_metadata <- Seurat_object@meta.data

sc_counts <- Seurat::GetAssayData(Seurat_object, assay = "RNA", slot = "counts")

# Only select cells that have a heterogeneity score (tumor cells)
sc_counts <- sc_counts[, !isNA(Seurat_metadata$heterogeneity_score)]

# Transpose matrix and convert to a dataframe to aggregate by sample
sc_counts_df <- as.data.frame(sparseToDenseMatrix(sc_counts %>% t()))

sc_counts_df[1:2, 1:2]

# Add annotation - tumor sample
sc_counts_df_sum <- sc_counts_df %>% rownames_to_column("sample") %>% left_join(
  Seurat_object@meta.data %>% filter(!isNA(heterogeneity_score)) %>% rownames_to_column("sample") %>% select(sample, annotation), by = "sample"
  ) %>% column_to_rownames("sample")

# Group by annotation and summarize all columns (except annotation)
sc_counts_df_sum <- sc_counts_df_sum %>% group_by(annotation) %>% summarize_at(colnames(sc_counts_df_sum)[!colnames(sc_counts_df_sum) %in% c("annotation")], sum)

sc_counts_df_sum[1:nrow(group_assignment), 1:5]

rm(sc_counts, sc_counts_df, Seurat_object)

gc()
```

```{r}
# transpose the dataframe and convert to a sparse matrix
sc_counts <- sc_counts_df_sum %>% column_to_rownames("annotation") %>% as.matrix() %>% t() %>% as("dgCMatrix")
rm(sc_counts_df_sum)

# Round counts to integers 
sc_counts <- round(sc_counts)
```

# DESeq2
```{r}
design_c <- group_assignment %>% select(group, CIN_status)
colnames(design_c) <- c("Sample", "CIN_status")

design_c %>% mutate_if(is.character, as.factor) %>% summary()
# head(sc_counts[, 1:2])

```

```{r}
# convert Sample and group columns to factors.
# Rows must be in the same order as the columns of the count matrix
df.diff.design <- design_c %>% arrange(Sample) %>% 
  mutate(
    Sample = as.factor(Sample),
    CIN_status = as.factor(CIN_status)
  )

head(df.diff.design)

as.data.frame(df.diff.design$Sample) %>% cbind(as.data.frame(colnames(sc_counts)))
```

Create the DESeq2 object and estimate size factors to apply counts normalization

```{r}
# create the deseq dataset
deseq.dataset <- DESeqDataSetFromMatrix(
  countData=sc_counts,
  colData=df.diff.design,
  design=~CIN_status
) %>%
  estimateSizeFactors()

# Change factor levels to compare CIN-High x CIN-low (default is alphabetical)
deseq.dataset$CIN_status <- factor(deseq.dataset$CIN_status, levels = c("CIN_low","CIN_high"))

```

## Dispersion estimation
```{r}
# create DESeq object
ddsWald <- DESeq(deseq.dataset, test = "Wald")

# Log fold shrinkage
ddsWald_apeglm <- lfcShrink(ddsWald, coef = "CIN_status_CIN_high_vs_CIN_low", type = "apeglm")

```

```{r}
resWald <- results(ddsWald)
summary(resWald)

df.deseq.pseudobulkWald <- resWald %>%
  as.data.frame() %>%
  rownames_to_column("gene")

df.deseq.pseudobulkWald.apeglm <- ddsWald_apeglm %>%
  as.data.frame() %>%
  rownames_to_column("gene")

# Export
df.deseq.pseudobulkWald %>% write.csv2(file.path(source.output.dir, "df.deseq.results.pseudobulkWald.csv"), row.names = FALSE)
 
df.deseq.pseudobulkWald.apeglm %>% write.csv2(file.path(source.output.dir, "df.deseq.results.pseudobulkWald.apeglm.csv"), row.names = FALSE)

```

## Heterogeneity score model
```{r}
design_c <- group_assignment %>% select(group, heterogeneity_score)
colnames(design_c) <- c("Sample", "heterogeneity_score")

design_c %>% mutate_if(is.character, as.factor) %>% summary()

```

```{r}
# convert Sample and group columns to factors.
# Rows must be in the same order as the columns of the count matrix
df.diff.design <- design_c %>% arrange(Sample) %>% 
  mutate(
    Sample = as.factor(Sample)
  )

# head(df.diff.design)
as.data.frame(df.diff.design$Sample) %>% cbind(as.data.frame(colnames(sc_counts)))

```

Create the DESeq2 object and estimate size factors to apply counts normalization

```{r}
# create the deseq dataset
deseq.score.dataset <- DESeqDataSetFromMatrix(
  countData=as.matrix(sc_counts),
  colData=df.diff.design,
  design=~heterogeneity_score
) %>%
  estimateSizeFactors()

```

```{r}
# create DESeq object
ddsWaldscore <- DESeq(deseq.score.dataset, test = "Wald")

ddsWaldscore_apeglm <- lfcShrink(ddsWaldscore, coef = "heterogeneity_score", type = "apeglm")
```

```{r}
resultsNames(ddsWaldscore)
resWaldscore <- results(ddsWaldscore)
summary(resWaldscore)

df.deseq.pseudobulk.Waldscore <- resWaldscore %>%
  as.data.frame() %>%
  rownames_to_column("gene")

df.deseq.pseudobulk.Waldscore.apeglm <- ddsWaldscore_apeglm %>%
  as.data.frame() %>%
  rownames_to_column("gene")

# Export
df.deseq.pseudobulk.Waldscore %>% write.csv2(file.path(source.output.dir, "df.deseq.results.pseudobulkWaldscore.csv"), row.names = FALSE)

df.deseq.pseudobulk.Waldscore.apeglm %>% write.csv2(file.path(source.output.dir, "df.deseq.results.pseudobulkWaldscore.apeglm.csv"), row.names = FALSE)
```

```{r}
CIN_signature <- c('PELI2','BMP2','SHH','TNS4','RAB3B','ROBO1','ARHGAP28','CHN2','CST1','F13A1','CPVL','SEMA6D','NHSL2','GTF2IP7','DPYSL3','PCDH7','KHDRBS3','TRAC','TMEM156', 'CST4','CD24','FGF5','NTN4')
```

# Volcano Plots
```{r}
alpha = 1e-3
log2FCthresh = 3

df.deseq.pseudobulkWald <- df.deseq.pseudobulkWald %>% mutate(
  within_threshold = (padj < alpha & abs(log2FoldChange) > log2FCthresh)
) %>% left_join(sc_means) 

df.deseq.pseudobulkWald.apeglm <- df.deseq.pseudobulkWald.apeglm %>% mutate(
  within_threshold = (padj < alpha & abs(log2FoldChange) > log2FCthresh)
) %>% left_join(sc_means) 

df.deseq.pseudobulkWald %>% ggplot(
  aes(x=log2FoldChange , y= -log10(pvalue + 1e-300), col = sc_mean, shape = pvalue > 0.01)) + 
  geom_point(alpha = 0.8) +
geom_vline(xintercept = 0, col = "black") + 
geom_vline(xintercept = c(-log2FCthresh, log2FCthresh), col = "red") +   geom_hline(yintercept = -log10(alpha), colour = "red") + 
geom_text_repel(
  aes(x = log2FoldChange, y = -log10(pvalue + 1e-300), label = gene), data = df.deseq.pseudobulkWald %>% filter(within_threshold), colour = "grey20", min.segment.length = 0
  ) +
  theme( plot.title = element_text(hjust= 0.5)) +
  labs(
    title = "Volcano plot Deseq2 Wald", 
    x = "Effect size: log2(fold-change)", 
    y = "-log10(p-value)"
  )  + scale_color_gradient2(low="red", mid = "gray", high="blue", midpoint = 1)

if(export){
  ggsave(file=file.path(source.output.dir,"deseq2Wald_pseudobulk_volcano.png"), width=6, height=4, dpi=300)
}

# Log fold shrunk apeglm
df.deseq.pseudobulkWald.apeglm  %>% ggplot(
  aes(x=log2FoldChange , y= -log10(pvalue + 1e-300), col = sc_mean, shape = pvalue > 0.01)) + 
  geom_point(alpha = 0.8) +
geom_vline(xintercept = 0, col = "black") + 
geom_vline(xintercept = c(-log2FCthresh, log2FCthresh), col = "red") +   geom_hline(yintercept = -log10(alpha), colour = "red") + 
geom_text_repel(
  aes(x = log2FoldChange, y = -log10(pvalue + 1e-300), label = gene), data = df.deseq.pseudobulkWald.apeglm %>% filter(within_threshold), colour = "grey20", min.segment.length = 0
  ) +
  theme(plot.title = element_text(hjust= 0.5)) +
  labs(
    title = "Volcano plot Deseq2 Wald apeglm", 
    x = "Effect size: log2(fold-change)", 
    y = "-log10(p-value)"
  )  + scale_color_gradient2(low="red", mid = "gray", high="blue", midpoint = 1)

if(export){
  ggsave(file=file.path(source.output.dir,"deseq2Waldapeglm_pseudobulk_volcano.png"), width=6, height=4, dpi=300)
}
```

```{r}
# Heterogeneity score model
df.deseq.pseudobulk.Waldscore <- df.deseq.pseudobulk.Waldscore %>% mutate(
  within_threshold = (padj < alpha & abs(log2FoldChange) > log2FCthresh)
) %>% left_join(sc_means) 

df.deseq.pseudobulk.Waldscore %>% ggplot(
  aes(x=log2FoldChange , y= -log10(pvalue + 1e-300), col = sc_mean)) +   geom_point(alpha = 0.8) +
geom_vline(xintercept = 0, col = "black") + 
geom_vline(xintercept = c(-log2FCthresh, log2FCthresh), col = "red") +   geom_hline(yintercept = -log10(alpha), colour = "red") + 
geom_text_repel(
  aes(x = log2FoldChange, y = -log10(pvalue + 1e-300), label = gene), data = df.deseq.pseudobulk.Waldscore %>% filter(gene %in% CIN_signature), colour = "grey20", min.segment.length = 0
  ) +
  theme(plot.title = element_text(hjust= 0.5)) +
  labs(
    title = "Volcano plot Heterogeneity Score Deseq2", 
    x = "Effect size: log2(fold-change)", 
    y = "-log10(p-value)"
  )  + scale_color_gradient2(low="red", mid = "gray", high="blue", midpoint = 1)

if(export){
  ggsave(file=file.path(source.output.dir,"deseq2Waldscore_pseudobulk_volcano.png"), width=6, height=4, dpi=300)
}


```

# Load msigdb hallmarks gene set
```{r Load msigdb hallmarks}
msigdb_hallmarks_set <- read.csv2(file = params$gene_sets)
msigdb_hallmarks_set <- split(msigdb_hallmarks_set$gene_symbol, msigdb_hallmarks_set$gs_name)

msigdb_hallmarks_names <- read.csv2(file = "names_hallmarks.csv")

```

# CIN High x CIN Low
## Run fGSEA
```{r}
# Using log2 fold change as a ranking metric
CIN_fc_rnk.deseq <- df.deseq.pseudobulkWald %>% select(gene, log2FoldChange) %>% deframe()
 
# fgseaRes.deseqstat <- read.csv2(file = file.path(source.output.dir, "scRNA_fgsea_CIN_highlow.csv"))

fgseaRes.deseq <- fgsea(pathways = msigdb_hallmarks_set,
                  stats    = CIN_fc_rnk.deseq,
                  minSize  = 15,
                  maxSize  = 500)

fgseaRes.deseq <- fgseaRes.deseq %>% mutate(within_threshold = padj < 0.01, "Enriched_in" = ifelse(NES > 0, "CIN-high", "CIN-low")) %>% left_join(msigdb_hallmarks_names)

# plotEnrichment(msigdb_hallmarks_set[["HALLMARK_INFLAMMATORY_RESPONSE"]],
#                CIN_fc_rnk) + labs(title="Inflammatory Response")

fgseaRes.deseq %>% select(NES, pval, padj) %>% summary()

# Plot results
fgseaRes.deseq %>% mutate(padj = ifelse(padj < 0.001, 0.00101, padj)) %>% ggplot(aes(x = log10(padj), y = NES)) +
  geom_point(aes(size = 0.5, alpha = 0.3, col = Enriched_in)) + 
  geom_hline(yintercept = 0, col = "black") +
  geom_text_repel(
  aes(x = log10(padj), y = NES, label = name), data = fgseaRes.deseq  %>% mutate(padj = ifelse(padj < 0.001, 0.00101, padj)) %>% filter(within_threshold | name %in% c("EMT", "Inflammatory Response", "IFN-a Response", "IFN-y Response", "TNF-a Signaling via NKFB", "IL2-STAT5 Signaling", "KRAS Signaling Up")), colour = "grey20", force = 3, force_pull = 2, min.segment.length = 0, max.overlaps = 20
  ) +
  labs(x = "FDRq", y = "Normalized Enrichment Score", col = "Enriched in") + 
  scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) + 
  scale_x_continuous(trans = "reverse", limits = c(0, -3), breaks = c(-3, -2, -1, 0) , labels = c("<0.001", "0.01", "0.1", "1")) + scale_fill_manual(values = c("red","blue")) + guides(alpha = FALSE, size = FALSE)

leading_edge <- fgseaRes.deseq %>% select(pathway, leadingEdge)

leading_edge_df <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("pathway", "leadingedge_gene", "position"))


for (i in 1:length(leading_edge$leadingEdge)){
  pathway_leadingedge <- leading_edge$leadingEdge[[i]] %>% as.data.frame() %>% rename("leadingedge_gene" = ".") %>% rownames_to_column("position") %>% mutate(pathway = leading_edge$pathway[i]) %>% relocate(pathway, 1) %>% relocate(position, .after = leadingedge_gene)
  
  leading_edge_df <- leading_edge_df %>% rbind(pathway_leadingedge)
  rm(i, pathway_leadingedge)
}

if (export){
ggsave(file = file.path(source.output.dir, "fGSEA_pseudobulk_deseq.png"), width=8, height=5, dpi=300)
}

if(export){
  fgseaRes.deseq %>% select(-leadingEdge) %>% write.csv2(file = file.path(source.output.dir, "pseudobulk_fgsea_deseq.csv"), row.names = FALSE)

  leading_edge_df %>% write.csv2(file = file.path(source.output.dir, "pseudobulk_fgsea_deseq_leadingedge.csv"), row.names = FALSE)
  }
```

## FGSEA DESeq Stat
```{r}
# Using Wald stat as a ranking metric
CIN_fc_rnk.deseqstat <- df.deseq.pseudobulkWald %>% select(gene, stat) %>% deframe()
 
# fgseaRes.deseqstat <- read.csv2(file = file.path(source.output.dir, "scRNA_fgsea_CIN_highlow.csv"))

fgseaRes.deseqstat <- fgsea(pathways = msigdb_hallmarks_set,
                  stats    = CIN_fc_rnk.deseqstat,
                  minSize  = 15,
                  maxSize  = 500)

fgseaRes.deseqstat <- fgseaRes.deseqstat %>% mutate(within_threshold = padj < 0.01, "Enriched_in" = ifelse(NES > 0, "CIN-high", "CIN-low")) %>% left_join(msigdb_hallmarks_names)

# plotEnrichment(msigdb_hallmarks_set[["HALLMARK_INFLAMMATORY_RESPONSE"]],
#                CIN_fc_rnk) + labs(title="Inflammatory Response")

fgseaRes.deseqstat %>% select(NES, pval, padj) %>% summary()

# Plot results
fgseaRes.deseqstat %>% mutate(padj = ifelse(padj < 0.001, 0.00101, padj)) %>% ggplot(aes(x = log10(padj), y = NES)) +
  geom_point(aes(size = 0.5, alpha = 0.3, col = Enriched_in)) + 
  geom_hline(yintercept = 0, col = "black") +
  geom_text_repel(
  aes(x = log10(padj), y = NES, label = name), data = fgseaRes.deseqstat  %>% mutate(padj = ifelse(padj < 0.001, 0.00101, padj)) %>% filter(within_threshold | name %in% c("EMT", "Inflammatory Response", "IFN-a Response", "IFN-y Response", "TNF-a Signaling via NKFB", "IL2-STAT5 Signaling", "KRAS Signaling Up")), colour = "grey20", force = 3, force_pull = 2, min.segment.length = 0
  ) +
  labs(x = "FDRq", y = "Normalized Enrichment Score", col = "Enriched in") + 
  scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) + 
  scale_x_continuous(trans = "reverse", limits = c(0, -3), breaks = c(-3, -2, -1, 0) , labels = c("<0.001", "0.01", "0.1", "1")) + scale_fill_manual(values = c("red","blue")) + guides(alpha = FALSE, size = FALSE)

leading_edge <- fgseaRes.deseqstat %>% select(pathway, leadingEdge)

leading_edge_df <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("pathway", "leadingedge_gene", "position"))


for (i in 1:length(leading_edge$leadingEdge)){
  pathway_leadingedge <- leading_edge$leadingEdge[[i]] %>% as.data.frame() %>% rename("leadingedge_gene" = ".") %>% rownames_to_column("position") %>% mutate(pathway = leading_edge$pathway[i]) %>% relocate(pathway, 1) %>% relocate(position, .after = leadingedge_gene)
  
  leading_edge_df <- leading_edge_df %>% rbind(pathway_leadingedge)
  rm(i, pathway_leadingedge)
}

if (export){
ggsave(file = file.path(source.output.dir, "fGSEA_pseudobulk_deseqstat.png"), width=8, height=5, dpi=300)
}

if(export){
  fgseaRes.deseqstat %>% select(-leadingEdge) %>% write.csv2(file = file.path(source.output.dir, "pseudobulk_fgsea_deseqstat.csv"), row.names = FALSE)

  leading_edge_df %>% write.csv2(file = file.path(source.output.dir, "pseudobulk_fgsea_deseqstat_leadingedge.csv"), row.names = FALSE)
  }
```

## FGSEA DESeq Score
```{r}
# Wald stat
CIN_fc_rnk.deseqstatscore <- df.deseq.pseudobulk.Waldscore %>% select(gene, stat) %>% deframe()
 
# fgseaRes.deseqstat <- read.csv2(file = file.path(source.output.dir, "scRNA_fgsea_CIN_highlow.csv"))

fgseaRes.scoredeseqstat <- fgsea(pathways = msigdb_hallmarks_set,
                  stats    = CIN_fc_rnk.deseqstatscore,
                  minSize  = 15,
                  maxSize  = 500)

fgseaRes.scoredeseqstat <- fgseaRes.scoredeseqstat %>% mutate(within_threshold = padj < 0.01, "Enriched_in" = ifelse(NES > 0, "CIN-high", "CIN-low")) %>% left_join(msigdb_hallmarks_names)

# plotEnrichment(msigdb_hallmarks_set[["HALLMARK_INFLAMMATORY_RESPONSE"]],
#                CIN_fc_rnk) + labs(title="Inflammatory Response")

fgseaRes.scoredeseqstat %>% select(NES, padj) %>% summary()

# Plot results
fgseaRes.scoredeseqstat %>% mutate(padj = ifelse(padj < 0.0001, 0.000101, padj)) %>% ggplot(aes(x = log10(padj), y = NES)) +
  geom_point(aes(size = 0.5, alpha = 0.3, col = Enriched_in)) + 
  geom_hline(yintercept = 0, col = "black") +
  geom_text_repel(
  aes(x = log10(padj), y = NES, label = name), data = fgseaRes.scoredeseqstat  %>% mutate(padj = ifelse(padj < 0.0001, 0.000101, padj)) %>% filter(within_threshold | name %in% c("EMT", "Inflammatory Response", "IFN-a Response", "IFN-y Response", "TNF-a Signaling via NKFB", "IL2-STAT5 Signaling", "KRAS Signaling Up")), colour = "grey20", force = 3, force_pull = 2, min.segment.length = 0, max.overlaps = 20
  ) +
  labs(x = "FDRq", y = "Normalized Enrichment Score", col = "Enriched in") + 
  scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) + 
  scale_x_continuous(trans = "reverse", limits = c(0, -4), breaks = c(-4, -3, -2, -1, 0) , labels = c("<0.0001", "0.001", "0.01", "0.1", "1")) + scale_fill_manual(values = c("red","blue")) + guides(alpha = FALSE, size = FALSE)

leading_edge <- fgseaRes.scoredeseqstat %>% select(pathway, leadingEdge)

leading_edge_df <- setNames(data.frame(matrix(ncol = 3, nrow = 0)), c("pathway", "leadingedge_gene", "position"))


for (i in 1:length(leading_edge$leadingEdge)){
  pathway_leadingedge <- leading_edge$leadingEdge[[i]] %>% as.data.frame() %>% rename("leadingedge_gene" = ".") %>% rownames_to_column("position") %>% mutate(pathway = leading_edge$pathway[i]) %>% relocate(pathway, 1) %>% relocate(position, .after = leadingedge_gene)
  
  leading_edge_df <- leading_edge_df %>% rbind(pathway_leadingedge)
  rm(i, pathway_leadingedge)
}

if (export){
ggsave(file = file.path(source.output.dir, "fGSEA_pseudobulk_scoredeseqstat.png"), width=8, height=5, dpi=300)
}

if(export){
  fgseaRes.scoredeseqstat %>% select(-leadingEdge) %>% write.csv2(file = file.path(source.output.dir, "pseudobulk_fgsea_scoredeseqstat.csv"), row.names = FALSE)

  leading_edge_df %>% write.csv2(file = file.path(source.output.dir, "pseudobulk_fgsea_scoredeseqstat_leadingedge.csv"), row.names = FALSE)
  }
```
# sessionInfo - Package Versions
```{r}
sessionInfo()
```