---
title: "Bakhoum et al 2018 scRNA-seq"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
---
# Loading Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning=FALSE, message = FALSE, dpi = 300
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# install.packages("BiocManager") #(only if you do not have BiocManager)
#BiocManager::install("DropletUtils")
# library(DropletUtils)
# install.packages("Seurat")
library(Seurat)
# install.packages("patchwork")
library(patchwork)
# install.packages("ggrepel")
library(ggrepel)
# install.packages("sparseMatrixStats")
library(sparseMatrixStats)
#install.packages("msigdbr")
library(msigdbr)
# BiocManager::install("GO.db")
library(GO.db)
# BiocManager::install("GOstats")
library(GOstats)
# BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)
# install.packages("tidyverse")
library(tidyverse)
# install.packages("proxy")
library(proxy)
# BiocManager::install("fgsea")
library(fgsea)
# install.packages('plotly')
library(plotly)
# install.packages('pheatmap')
library(pheatmap)
# install.packages("gplots")
library(gplots)

select = dplyr::select
rename = dplyr::rename
```

# Loading data
```{r}
# Load filtered and combined data
scRNA_c <- readRDS(file = "data/scRNA_c.rds")

export = FALSE

# generate output dir path named data
source.output.dir <- file.path("output_SCT")

# if source output dir does not exist, create it
if (!dir.exists(source.output.dir)) {
    dir.create(source.output.dir)
} else{
  print("Output folder already exists")
}

```

## Load msigdb hallmarks gene set
```{r}
all_gene_sets <- msigdbr("Homo sapiens")

msigdb_hallmarks_set <- filter(all_gene_sets, gs_cat == "H") %>% select(gs_name, gene_symbol)

msigdb_hallmarks_set <- split(msigdb_hallmarks_set$gene_symbol, msigdb_hallmarks_set$gs_name)

msigdb_hallmarks_names <- read.csv2("data/names_hallmarks.csv")

rm(all_gene_sets)
# Another option is loading files with gene sets
# msigdb_hallmarks <- gmtPathways("data/h.all.v7.4.symbols.gmt")
# msigdb_hallmarks %>% head() %>% lapply(head)

```

# Using SCTransform

```{r}
scRNA_SCT <- SCTransform(scRNA_c, verbose = FALSE)

# Add CIN status meta data
scRNA_SCT@meta.data <- scRNA_SCT@meta.data %>% mutate(CIN_status = ifelse(orig.ident %in% c("dnMCAK"),"CIN-high","CIN-low"))

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Score cell cycle
scRNA_SCT <- CellCycleScoring(scRNA_SCT, s.features = s.genes, g2m.features = g2m.genes)

```

## Perform PCA
```{r}
# Perform PCA
scRNA_SCT <- RunPCA(scRNA_SCT, npcs = 200)

# Plot PCA results
Idents(scRNA_SCT) <- "orig.ident"
VizDimLoadings(scRNA_SCT, dims = 1:2, reduction = "pca")
DimPlot(scRNA_SCT, reduction = "pca")
ElbowPlot(scRNA_SCT, ndims = 200)

# Check the total variance explained
# On Seurat 3:
pca <- scRNA_SCT[["pca"]]
# Get the total variance of the scaled counts matrix:
mat <- Seurat::GetAssayData(scRNA_SCT, assay = "SCT", slot = "scale.data")

total_variance <- sum(matrixStats::rowVars(mat))

eigValues = (pca@stdev)^2  ## EigenValues
varExplained = eigValues / total_variance

varExplained

cumsum(varExplained) %>% as.data.frame() %>% rename("cumvar" = ".") %>% ggplot(aes(x = c(1:200), y = cumvar)) + geom_point() + labs(x = "PC #", y = "Cumulative Variance")

rm(mat)
rm(pca)

```

## Run UMAP and TSNE
```{r}
# This is not using the cell cycle regressed data! the visualization looks the same as before

# Dimensionality reduction with UMAP on variable features instead of PCs
scRNA_SCT <- RunUMAP(scRNA_SCT, reduction.use = NULL, features = VariableFeatures(scRNA_SCT))

# Store this UMAP as a custom reduction
scRNA_SCT[["UMAP_features"]] <- CreateDimReducObject(embeddings = scRNA_SCT@reductions$umap@cell.embeddings, key = "UMAP_features", assay = DefaultAssay(scRNA_SCT))

# Run UMAP and project in 3D dimensions
scRNA_SCT <- RunUMAP(scRNA_SCT, dims = 1:200, n.components = 3L)

# Store this UMAP as a custom reduction
scRNA_SCT[["umap3d"]] <- CreateDimReducObject(embeddings = scRNA_SCT@reductions$umap@cell.embeddings, key = "umap3d", assay = DefaultAssay(scRNA_SCT))


# PCs seem to be using the regressed data as intended

# Dimensionality reduction with UMAP on PCs
scRNA_SCT <- RunUMAP(scRNA_SCT, dims = 1:200)

```

## Clustering cells
```{r}
# Clustering cells
scRNA_SCT <- FindNeighbors(scRNA_SCT, dims = 1:200)
scRNA_SCT <- FindClusters(scRNA_SCT, resolution = 0.5)

scRNA_SCT <- StashIdent(scRNA_SCT, save.name = "clusterID")

# Inspect cluster information
#head(Idents(scRNA_cn), 5)
Idents(scRNA_SCT) <- "clusterID"

table(scRNA_SCT$clusterID)
prop.table(table(scRNA_SCT$clusterID))
prop.table(table(scRNA_SCT$clusterID, scRNA_SCT$orig.ident), margin = 2)
prop.table(table(scRNA_SCT$orig.ident, scRNA_SCT$clusterID), margin = 2)

prop.table(table(scRNA_SCT$clusterID, scRNA_SCT$CIN_status), margin = 2)
prop.table(table(scRNA_SCT$CIN_status, scRNA_SCT$clusterID), margin = 2)

table(scRNA_SCT$clusterID, scRNA_SCT$CIN_status) %>% as.data.frame() %>% rename("clusterID" = Var1, "group" = Var2) %>% ggplot(aes(x = clusterID, y = group, fill = Freq)) + geom_tile() + scale_fill_distiller(palette = "PuBu") + geom_text(aes(label=Freq)) + labs(y = "", fill = "Number of cells")

```
## Visualization
```{r}
DimPlot(scRNA_SCT, reduction = "UMAP_features", group.by = "CIN_status")
DimPlot(scRNA_SCT, reduction = "UMAP_features", group.by = "orig.ident")
DimPlot(scRNA_SCT, reduction = "UMAP_features", group.by = "Phase")
DimPlot(scRNA_SCT, reduction = "UMAP_features", group.by = "clusterID")

DimPlot(scRNA_SCT, reduction = "umap", group.by = "CIN_status")
DimPlot(scRNA_SCT, reduction = "umap", group.by = "orig.ident")
DimPlot(scRNA_SCT, reduction = "umap", group.by = "Phase")
DimPlot(scRNA_SCT, reduction = "umap", group.by = "clusterID")

```

## 3D UMAP
```{r}
# Visualize what headings are called so that you can extract them to form a dataframe
Embeddings(object = scRNA_SCT, reduction = "umap3d") %>% head()

# Prepare a dataframe for cell plotting
plot.data <- FetchData(object = scRNA_SCT, vars = c("umap3d_1", "umap3d_2", "umap3d_3", "seurat_clusters", "Phase", "orig.ident"))

# Make a column of row name identities (these will be your cell/barcode names)
plot.data$label <- paste(rownames(plot.data))

# Plot your data
#When you visualize your plotly object, hovering your mouse pointer over a point shows cell names

fig <- plot_ly(data = plot.data, 
        x = ~umap3d_1, y = ~umap3d_2, z = ~umap3d_3, 
        color = ~seurat_clusters, 
        colors = c("lightseagreen", "gray50", "darkgreen", "red4", "red", "turquoise4", "black", "yellow4", "royalblue1", "lightcyan3", "peachpuff3",
"khaki3", "gray20", "orange2", "royalblue4", "yellow3", "gray80", "darkorchid1", "lawngreen", "plum2", "darkmagenta"),
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 3, width=2), # controls size of points
        text=~orig.ident,
        hoverinfo="text") 
fig

fig <- plot_ly(data = plot.data, 
        x = ~umap3d_1, y = ~umap3d_2, z = ~umap3d_3, 
        color = ~Phase,
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 3, width=2), # controls size of points
        text=~orig.ident,
        hoverinfo="text") 

fig

rm(plot.data)

```

## Export
```{r}
if(export){
saveRDS(scRNA_SCT, file = paste0(source.output.dir, "/Bakhoum_scRNA_SCT.rds"))

}

scRNA_SCT <- readRDS(file = paste0(source.output.dir, "/Bakhoum_scRNA_SCT.rds"))

```

# Using SCTransform and regressing cell cycle

```{r}
scRNA_SCT_cellreg <- SCTransform(scRNA_c, verbose = FALSE)

# Add CIN status meta data
scRNA_SCT_cellreg@meta.data <- scRNA_SCT_cellreg@meta.data %>% mutate(CIN_status = ifelse(orig.ident %in% c("dnMCAK"),"CIN-high","CIN-low"))

# Score cell cycle
scRNA_SCT_cellreg <- CellCycleScoring(scRNA_SCT_cellreg, s.features = s.genes, g2m.features = g2m.genes)

# Regress cell cycle scores
vars_to_regress <- c("S.Score", "G2M.Score")

scRNA_SCT_cellreg <- SCTransform(scRNA_SCT_cellreg, vars.to.regress = vars_to_regress, verbose = FALSE)

```
## Perform PCA
```{r}
# Perform PCA
scRNA_SCT_cellreg <- RunPCA(scRNA_SCT_cellreg, npcs = 200)

# Plot PCA results
Idents(scRNA_SCT_cellreg) <- "orig.ident"
VizDimLoadings(scRNA_SCT_cellreg, dims = 1:2, reduction = "pca")
DimPlot(scRNA_SCT_cellreg, reduction = "pca")
ElbowPlot(scRNA_SCT_cellreg, ndims = 200)

# Check the total variance explained
# On Seurat 3:
pca <- scRNA_SCT_cellreg[["pca"]]
# Get the total variance of the scaled counts matrix:
mat <- Seurat::GetAssayData(scRNA_SCT_cellreg, assay = "SCT", slot = "scale.data")

total_variance <- sum(matrixStats::rowVars(mat))

eigValues = (pca@stdev)^2  ## EigenValues
varExplained = eigValues / total_variance

varExplained

cumsum(varExplained) %>% as.data.frame() %>% rename("cumvar" = ".") %>% ggplot(aes(x = c(1:200), y = cumvar)) + geom_point() + labs(x = "PC #", y = "Cumulative Variance")

rm(mat)
rm(pca)

```

## Clustering cells
```{r}
# Clustering cells
scRNA_SCT_cellreg <- FindNeighbors(scRNA_SCT_cellreg, dims = 1:200)
scRNA_SCT_cellreg <- FindClusters(scRNA_SCT_cellreg, resolution = 0.5)

scRNA_SCT_cellreg <- StashIdent(scRNA_SCT_cellreg, save.name = "clusterID")

# Inspect cluster information
#head(Idents(scRNA_cn), 5)
Idents(scRNA_SCT_cellreg) <- "clusterID"

table(scRNA_SCT_cellreg$clusterID)
prop.table(table(scRNA_SCT_cellreg$clusterID))
prop.table(table(scRNA_SCT_cellreg$clusterID, scRNA_SCT_cellreg$orig.ident), margin = 2)
prop.table(table(scRNA_SCT_cellreg$orig.ident, scRNA_SCT_cellreg$clusterID), margin = 2)

prop.table(table(scRNA_SCT_cellreg$clusterID, scRNA_SCT_cellreg$CIN_status), margin = 2)
prop.table(table(scRNA_SCT_cellreg$CIN_status, scRNA_SCT_cellreg$clusterID), margin = 2)

table(scRNA_SCT$clusterID, scRNA_SCT_cellreg$CIN_status) %>% as.data.frame() %>% rename("clusterID" = Var1, "group" = Var2) %>% ggplot(aes(x = clusterID, y = group, fill = Freq)) + geom_tile() + scale_fill_distiller(palette = "PuBu") + geom_text(aes(label=Freq)) + labs(y = "", fill = "Number of cells")

```

```{r}

prop.table(table(scRNA_SCT_cellreg$CIN_status, scRNA_SCT_cellreg$clusterID), margin = 2)

cluster_assignment <- prop.table(table(scRNA_SCT_cellreg$CIN_status, scRNA_SCT_cellreg$clusterID), margin = 2) %>% as.data.frame() %>% rename("CIN_status" = Var1, "cluster" = Var2)

cluster_assignment$CIN_status <- gsub("-", "_", cluster_assignment$CIN_status)

cluster_assignment <- cluster_assignment %>% pivot_wider(values_from = Freq, names_from = CIN_status) %>% mutate(CIN_status = ifelse(CIN_high < 0.15, "CIN-low", "CIN-high"))

cluster_assignment
```

## Run UMAP and TSNE
```{r}
# Dimensionality reduction with UMAP on variable features instead of PCs
scRNA_SCT_cellreg <- RunUMAP(scRNA_SCT_cellreg, assay = "SCT", reduction.use = NULL, features = VariableFeatures(scRNA_SCT_cellreg))

# Store this UMAP as a custom reduction
scRNA_SCT_cellreg[["UMAP_features"]] <- CreateDimReducObject(embeddings = scRNA_SCT_cellreg@reductions$umap@cell.embeddings, key = "UMAP_features", assay = DefaultAssay(scRNA_SCT_cellreg))

# Run UMAP and project in 3D dimensions
scRNA_SCT_cellreg <- RunUMAP(scRNA_SCT_cellreg, dims = 1:200, n.components = 3L)

# Store this UMAP as a custom reduction
scRNA_SCT_cellreg[["umap3d"]] <- CreateDimReducObject(embeddings = scRNA_SCT_cellreg@reductions$umap@cell.embeddings, key = "umap3d", assay = DefaultAssay(scRNA_SCT_cellreg))

# PCs seem to be using the regressed data as intended

# Dimensionality reduction with UMAP on PCs
scRNA_SCT_cellreg <- RunUMAP(scRNA_SCT_cellreg, dims = 1:200)

```

## Visualization
```{r}
DimPlot(scRNA_SCT_cellreg, reduction = "umap", group.by = "CIN_status")
DimPlot(scRNA_SCT_cellreg, reduction = "umap", group.by = "orig.ident")
DimPlot(scRNA_SCT_cellreg, reduction = "umap", group.by = "Phase")
DimPlot(scRNA_SCT_cellreg, reduction = "umap", group.by = "clusterID")
```

## 3D UMAP
```{r}
# Visualize what headings are called so that you can extract them to form a dataframe
Embeddings(object = scRNA_SCT_cellreg, reduction = "umap3d") %>% head()

# Prepare a dataframe for cell plotting
plot.data <- FetchData(object = scRNA_SCT_cellreg, vars = c("umap3d_1", "umap3d_2", "umap3d_3", "seurat_clusters", "Phase", "orig.ident"))

# Make a column of row name identities (these will be your cell/barcode names)
plot.data$label <- paste(rownames(plot.data))

# Plot your data
#When you visualize your plotly object, hovering your mouse pointer over a point shows cell names

fig <- plot_ly(data = plot.data, 
        x = ~umap3d_1, y = ~umap3d_2, z = ~umap3d_3, 
        color = ~seurat_clusters, 
        colors = c("lightseagreen", "gray50", "darkgreen", "red4", "red", "turquoise4",
                   "black", "yellow4", "royalblue1", "lightcyan3", "peachpuff3",
                   "khaki3", "gray20", "orange2", "royalblue4", "yellow3", "gray80",
                   "darkorchid1", "lawngreen", "plum2", "darkmagenta"),
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 3, width=2), # controls size of points
        text=~orig.ident,
        hoverinfo="text") 
fig

fig <- plot_ly(data = plot.data, 
        x = ~umap3d_1, y = ~umap3d_2, z = ~umap3d_3, 
        color = ~Phase,
        type = "scatter3d", 
        mode = "markers", 
        marker = list(size = 3, width=2), # controls size of points
        text=~orig.ident,
        hoverinfo="text") 

fig

rm(plot.data)

```

## Cluster markers
```{r}
# Find cluster markers 
Idents(scRNA_SCT_cellreg) <- "clusterID"
SCT_cellreg.cluster.DE <- FindAllMarkers(scRNA_SCT_cellreg, min.pct = 0.05, logfc.threshold = 0)

#cluster.markers
clustertop10 <- SCT_cellreg.cluster.DE %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

SCT_cellreg.cluster.DE %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)

```

## Plotting signatures
```{r}
Idents(scRNA_SCT_cellreg) <- "clusterID"

# Plotting EMT signature 
EMT_signature <- c("VIM", "ZEB2", "SNAI2", "ZEB1")

VlnPlot(scRNA_SCT_cellreg, features = EMT_signature)
FeaturePlot(scRNA_SCT_cellreg, features = EMT_signature)

EMT_signature2 <- c("EZH", "JUN", "VIM", "STEAP1", "SOX4", "MMP14", "SHH", "TIMP1", "ZEB1")

EMT_signature2 <- EMT_signature2[EMT_signature2 %in% rownames(scRNA_SCT_cellreg@assays$RNA)]

Idents(scRNA_SCT_cellreg) <- "clusterID"
for(i in 1:length(EMT_signature2)){
p <- VlnPlot(scRNA_SCT_cellreg, features = EMT_signature2[i])
print(p)
}

```

```{r}
# Plotting Genes enriched in the "M" Population from Bakhoum et al
M_markers = c("ITGB5", "ITGB1", "ITGA5", "ITGA10", "IGFBP4", "FN1", "DSC2", "CXCL1","CTNNB1", "BMP4", "BCL2L1")

for(i in 1:length(M_markers)){
p <- VlnPlot(scRNA_SCT_cellreg, features = M_markers[i])
print(p)
}
```

```{r}
# Plot CIN signature from Bakhoum et al 2018 (Supp. table 5), comparing CIN high and CIN low
CIN_signature <- c('PELI2','BMP2','SHH','TNS4','RAB3B','ROBO1','ARHGAP28','CHN2','CST1','F13A1','CPVL','SEMA6D','NHSL2','GTF2IP7','DPYSL3','PCDH7','KHDRBS3','TRAC','TMEM156',
'CST4','CD24','FGF5','NTN4')

CIN_present <- CIN_signature %in% rownames(scRNA_SCT_cellreg@assays$RNA)

for(i in 1:length(CIN_present)){
p <- VlnPlot(scRNA_SCT_cellreg, features = CIN_signature[CIN_present][i])
print(p)
}

```

## FGSEA DE between clusters
```{r}
scRNA_SCT_cellreg.l <- SCT_cellreg.cluster.DE %>% select(gene, cluster, avg_log2FC) %>% split(f = SCT_cellreg.cluster.DE$cluster)

SCT.cellreg.cluster.DE.fgsea <- scRNA_SCT_cellreg.l %>% lapply(FUN = function(x){
  x <- x %>% select(-cluster) %>% deframe()
  x <- fgsea(pathways = msigdb_hallmarks_set, 
                  stats = x,
                  minSize = 15,
                  maxSize = 500)
})


# Plot results for the GSEA on each cluster 
lapply(seq_along(SCT.cellreg.cluster.DE.fgsea), FUN = function(i){
  SCT.cellreg.cluster.DE.fgsea[[i]] <- SCT.cellreg.cluster.DE.fgsea[[i]] %>% mutate(
    within_threshold = pval < 0.05, 
    pval = ifelse(pval < 0.0001, 0.000101, pval)
    ) %>% left_join(msigdb_hallmarks_names)

  SCT.cellreg.cluster.DE.fgsea[[i]] %>% ggplot(aes(x = log10(pval), y = NES)) +
  geom_point(aes(size = 0.5, alpha = 0.3, col = as.factor(sign(NES)))) + 
  geom_hline(yintercept = 0, col = "black") +
  geom_text_repel(
  aes(x = log10(pval), y = NES, label = name), data = SCT.cellreg.cluster.DE.fgsea[[i]] %>% filter(within_threshold), colour = "grey20", force = 3, force_pull = 2, min.segment.length = 0
  ) +
  labs(x = "log10(pval)", y = "Normalized Enrichment Score", col = "Enriched in Cluster", title = paste("Cluster: ", i - 1, sep = "")) + 
  scale_y_continuous(limits = c(-3,3), breaks = seq(-3,3,1)) + 
  scale_x_continuous(trans = "reverse", limits = c(0, -4), breaks = c(-4, -3, -2, -1, 0) , labels = c("<0.0001", "0.001", "0.01", "0.1", "1")) + scale_fill_manual(values = c("red","blue")) + guides(alpha = FALSE, size = FALSE)
})


# Combine summaries from the GO enrichment tests
SCT.cellreg.cluster.DE.fgsea_summary <- SCT.cellreg.cluster.DE.fgsea[[1]] %>% mutate(cluster = 0) %>% filter(pval == "nothing")

for(i in 1:length(SCT.cellreg.cluster.DE.fgsea)){
  print(i)
  SCT.cellreg.cluster.DE.fgsea_summary <- rbind(
    SCT.cellreg.cluster.DE.fgsea_summary, 
    SCT.cellreg.cluster.DE.fgsea[[i]] %>% mutate(cluster = i-1)
    )
}

if(export){
SCT.cellreg.cluster.DE.fgsea_summary %>% select(-leadingEdge) %>% write.csv2(file = paste0(source.output.dir, "/scRNA_SCT_cellreg_cluster_fgsea.csv"), row.names = FALSE)
  }

SCT.cellreg.cluster.DE.fgsea_summary <- read.csv2(file = paste0(source.output.dir, "/scRNA_SCT_cellreg_cluster_fgsea.csv"))
SCT.cellreg.cluster.DE.fgsea <- SCT.cellreg.cluster.DE.fgsea_summary %>% split(f = SCT.cellreg.cluster.DE.fgsea_summary$cluster)

```
## Heatmap Cluster NES
```{r}
colfunc <- colorRampPalette(c("blue", "white", "red"))

SCT.cellreg.cluster.DE.fgsea_summary %>% left_join(msigdb_hallmarks_names) %>% select(name, NES, cluster) %>% pivot_wider(values_from = NES, names_from = name) %>% column_to_rownames("cluster") %>% mutate_all(~replace(., is.na(.), 0)) %>% as.matrix() %>% pheatmap(col=colfunc(10))

SCT.cellreg.cluster.DE.fgsea_summary %>% left_join(msigdb_hallmarks_names) %>% select(name, NES, cluster) %>% pivot_wider(values_from = NES, names_from = name)  %>% 
  column_to_rownames(var = "cluster") %>% mutate_all(~replace(., is.na(.), 0)) %>% as.matrix() %>%
  heatmap.2(col=colfunc(10), dendrogram = "both", trace = "none", srtCol=75)

```
## Export
```{r}
if(export){
saveRDS(scRNA_SCT_cellreg, file = paste0(source.output.dir, "/Bakhoum_scRNA_SCT_cellreg.rds"))
  
SCT_cellreg.cluster.DE %>% write.csv2(file = paste0(source.output.dir, "/Bakhoum_scRNA_SCT_cellreg_cluster_markers.csv")
, row.names = FALSE)

}


scRNA_SCT_cellreg <- readRDS(file = paste0(source.output.dir, "/Bakhoum_scRNA_SCT_cellreg.rds"))
SCT_cellreg.cluster.DE <- read.csv2(file = paste0(source.output.dir, "/Bakhoum_scRNA_SCT_cellreg_cluster_markers.csv"))
```

#----------------------------------------------------------------------

# Inspect Batch effects
Look for batch effects by scaling and normalizing each dataset individually first, then merging
```{r}
# To compare batch effects, normalize and scale each dataset individually, then merge

# Kif2b
median_library_size <- kif2b_f@assays$RNA %>% colSums() %>% median()
# Normalize dataset
kif2b_fn <- NormalizeData(kif2b_f, normalization.method = "LogNormalize", scale.factor = median_library_size)
# find highly variable features with a variance stabilizing transform
kif2b_fn <- FindVariableFeatures(kif2b_fn, selection.method = "vst", nfeatures = 2000)


# Scale dataset
kif2b_fn <- ScaleData(kif2b_fn, features = rownames(kif2b_fn))

# MCAK
median_library_size <- MCAK_f@assays$RNA %>% colSums() %>% median()
# Normalize dataset
MCAK_fn <- NormalizeData(MCAK_f, normalization.method = "LogNormalize", scale.factor = median_library_size)
# find highly variable features with a variance stabilizing transform
MCAK_fn <- FindVariableFeatures(MCAK_fn, selection.method = "vst", nfeatures = 2000)
# Scale dataset
MCAK_fn <- ScaleData(MCAK_fn, features = rownames(MCAK_fn))


# dnMCAK
median_library_size <- dnMCAK_f@assays$RNA %>% colSums() %>% median()
# Normalize dataset
dnMCAK_fn <- NormalizeData(dnMCAK_f, normalization.method = "LogNormalize", scale.factor = median_library_size)
# find highly variable features with a variance stabilizing transform
dnMCAK_fn <- FindVariableFeatures(dnMCAK_fn, selection.method = "vst", nfeatures = 2000)
# Scale dataset
dnMCAK_fn <- ScaleData(dnMCAK_fn, features = rownames(dnMCAK_fn))


# Merge datasets, keeping the normalization and scaling on the data
scRNA_norm_c <- merge(kif2b_fn, y = c(MCAK_fn, dnMCAK_fn), add.cell.ids = c("kif2b", "MCAK", "dnMCAK"), project = "BakhoumCIN2018", merge.data = TRUE)

# Normalize, find variable features and scale the merged dataset
scRNA_norm_cn <- NormalizeData(scRNA_norm_c, normalization.method = "LogNormalize")
scRNA_norm_cn <- FindVariableFeatures(scRNA_norm_cn, selection.method = "vst", nfeatures = 2000)
scRNA_norm_cn <- ScaleData(scRNA_norm_cn, features = rownames(scRNA_norm_cn))

rm(scRNA_norm_c)
rm(kif2b_f)
rm(MCAK_f)
rm(dnMCAK_f)

```

```{r}
# Perform PCA
scRNA_norm_cn <- RunPCA(scRNA_norm_cn, features = VariableFeatures(object = scRNA_norm_cn), npcs = 200)

# Plot PCA results
DimPlot(scRNA_norm_cn, reduction = "pca", group.by = "orig.ident")
ElbowPlot(scRNA_norm_cn, ndims = 200)

# Cluster the cells using the first 200 principal components.

scRNA_norm_cn <- FindNeighbors(scRNA_norm_cn, dims = 1:200)
scRNA_norm_cn <- FindClusters(scRNA_norm_cn, resolution = 0.5)

scRNA_norm_cn <- StashIdent(scRNA_norm_cn, save.name = "clusterID")

scRNA_norm_cn <- RunUMAP(scRNA_norm_cn, dims = 1:200)
scRNA_norm_cn <- RunTSNE(scRNA_norm_cn, dims.use = 1:200, reduction.use = "pca", perplexity = 30)

```

```{r}
# Inspect dataset, looking for batch effects
VlnPlot(scRNA_norm_cn, features = c("nFeature_RNA", "nCount_RNA"), group.by = "orig.ident", ncol=2)

# Plot PCA results
DimPlot(scRNA_norm_cn, reduction = "pca", group.by = "orig.ident")

DimPlot(scRNA_norm_cn, reduction = "umap", group.by = "clusterID")
DimPlot(scRNA_norm_cn, reduction = "umap", group.by = "orig.ident")
DimPlot(scRNA_norm_cn, reduction = "tsne", group.by = "clusterID")
DimPlot(scRNA_norm_cn, reduction = "tsne", group.by = "orig.ident")

```

```{r}
# Export Seurat object with processed individual datasets prior to merging, re-normalizing and scaling again
if(export){
saveRDS(scRNA_norm_cn, file = "data/Bakhoum_scRNA_norm_cn.rds")
  }
#scRNA_cn <- readRDS(file = "data/Bakhoum_scRNA_norm_cn.rds")

rm(scRNA_norm_cn)
```

# Seurat integration
Try to correct for batch effects with Seurat integration 
```{r}
# Explore batch correction with Seurat integration workflow
ob.list <- list(kif2b_fn, MCAK_fn, dnMCAK_fn)

# ob.list <- lapply(list(kif2b_f, MCAK_f, dnMCAK_f), FUN = function(x){
#   x <- NormalizeData(x)
#   x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
# }
# )

features <- SelectIntegrationFeatures(object.list = ob.list)

CINanchors <- FindIntegrationAnchors(object.list = ob.list, anchor.features = features)

IntegratedCIN <- IntegrateData(anchorset = CINanchors)

# Assign and inspect cell cycle scores
IntegratedCIN <- CellCycleScoring(IntegratedCIN, s.features = s.genes, g2m.features = g2m.genes)

```

```{r}
# specify that we will perform downstream analysis on the corrected data note that the original
# unmodified data still resides in the 'RNA' assay
DefaultAssay(IntegratedCIN) <- "integrated"

# Run the standard workflow for visualization and clustering
IntegratedCIN <- ScaleData(IntegratedCIN, verbose = FALSE)
IntegratedCIN <- RunPCA(IntegratedCIN, npcs = 200, verbose = FALSE)
IntegratedCIN <- RunUMAP(IntegratedCIN, reduction = "pca", dims = 1:200)
IntegratedCIN <- RunTSNE(IntegratedCIN, reduction = "pca", dims = 1:200)
IntegratedCIN <- FindNeighbors(IntegratedCIN, reduction = "pca", dims = 1:200)
IntegratedCIN <- FindClusters(IntegratedCIN, resolution = 0.5)

DimPlot(IntegratedCIN, reduction = "umap", group.by = "orig.ident")
DimPlot(IntegratedCIN, reduction = "umap", group.by = "Phase")
DimPlot(IntegratedCIN, reduction = "umap", split.by = "orig.ident")

DimPlot(IntegratedCIN, reduction = "umap", label = TRUE, repel = TRUE)

DimPlot(IntegratedCIN, reduction = "tsne", group.by = "orig.ident")
DimPlot(IntegratedCIN, reduction = "tsne", group.by = "Phase")
DimPlot(IntegratedCIN, reduction = "tsne", label = TRUE, repel = TRUE)

```

```{r}
if(export){
saveRDS(IntegratedCIN, file = "data/Bakhoum_integrated_scRNA.rds")
}
```

