---
title: "Nelson et al 2020 scRNA-seq"
author: "Pedro Batista Tan"
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
output: 
  html_document:
    toc: true
    number_sections: true
params:
    export: TRUE

---
# Loading Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning=FALSE, message = FALSE, dpi = 300, error = TRUE
  # fig.width = 10, fig.height = 6, output: html_notebook
)

# install.packages("BiocManager") #(only if you do not have BiocManager)
# BiocManager::install("DropletUtils", "org.Hs.eg.db")
# install.packages("Seurat", "sparseMatrixStats", "tidyverse")
library(Seurat)
library(sparseMatrixStats)
library(tidyverse)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(plotly)

source(paste0(dirname(getwd()), "/Helper_functions.R"))

select = dplyr::select

```

# Setup
```{r Setup}
export = params$export

# generate output dir path named data
source.output.dir <- file.path("output_sc")

Seurat_object_name <- "scNelson"

# if source output dir does not exist, create it
if (!dir.exists(source.output.dir)) {
    dir.create(source.output.dir)
} else{
  print("Output folder already exists")
}

```

# Loading data
```{r Loading data}
barcodes <- read.table("data/E-MTAB-8559.aggregated_filtered_counts.mtx_cols")

genes <- read.table("data/E-MTAB-8559.aggregated_filtered_counts.mtx_rows")
# matrix <- read.table("data/E-MTAB-8410.aggregated_filtered_counts.mtx", sep = "\ ", header = TRUE, fill = TRUE)

matrix <- read.table("data/E-MTAB-8559.aggregated_filtered_counts.mtx", sep = "\ ", skip = 2)

design <- read.table("data/ExpDesign-E-MTAB-8559.tsv", sep = "\t", header = TRUE)

study_metadata = read.table(file = "data/E-MTAB-8559.sdrf.txt", header = TRUE, sep = "\t")

```
# Map ensembl IDs to gene symbols
```{r}
gene_features <- read.csv2(file = paste0(dirname(getwd()), "/biomaRt_gene_features.csv"))

gene_map <- genes %>% rename("ensembl_gene_id" = "V1") %>% left_join(gene_features %>% select(ensembl_gene_id, hgnc_symbol)) %>% unique()

gene_map %>% filter(is.na(hgnc_symbol)) %>% nrow()
gene_map %>% filter(hgnc_symbol == "") %>% nrow()

gene_map_f <- gene_map %>% filter(!is.na(hgnc_symbol)) %>% filter(hgnc_symbol != "")
```

```{r}
# Export features and barcodes to import with Read10x function
# write.table(genes, file = "data/data_10x_format/features.tsv", row.names=FALSE, sep="\t", col.names = FALSE, quote = FALSE)

# write.table(barcodes, file = "data/data_10x_format/barcodes.tsv", row.names=FALSE, sep="\t", col.names  = FALSE , quote = FALSE)

rm(barcodes, genes, matrix)
```

```{r}
# Get genes with duplicate symbols
duplicates <- table(gene_map_f$hgnc_symbol) %>% as.data.frame() %>% rename("hgnc_symbol" = "Var1") %>% filter(Freq > 1) %>% inner_join(gene_map_f)
```
```{r}
# data <- ReadMtx(mtx = expression_matrix, cells = barcodes, features = genes)
data_dir <- "data/data_10x_format"
list.files(data_dir) # Should show barcodes.tsv, features.tsv, and matrix.mtx
data <- Read10X(data.dir = data_dir)
```

```{r}
data_df <- as.data.frame(sparseToDenseMatrix(data))

data_df$hgnc_symbol <- gene_map$hgnc_symbol

head(data_df[,20318:20322])

rm(data)

```
# Aggregate counts for duplicate symbols
```{r}
# Summarize
data_df_duplicates = data_df[rownames(data_df) %in% duplicates$ensembl_gene_id, ]

data_sum <- data_df_duplicates %>% group_by(hgnc_symbol) %>% summarize_at(colnames(data_df_duplicates)[!colnames(data_df_duplicates) %in% c("hgnc_symbol")], sum)
  
rm(data_df_duplicates)
head(data_sum)
```

```{r}
# Join summarized dataset
rownames(data_df) <- c()
rownames(data_sum) <- c()

# discard genes without symbols
data_df <- data_df %>% filter(hgnc_symbol %in% gene_map_f$hgnc_symbol)

# Remove duplicated genes
data_df <- data_df[!data_df$hgnc_symbol %in% duplicates$hgnc_symbol,]

# Add sum of duplicated genes
rownames(data_df) <- c()
data_df <- data_df %>% rbind(data_sum) %>% column_to_rownames("hgnc_symbol")

data = as.matrix(data_df)

rm(data_df, data_sum)

data = as(data, "dgCMatrix")

```

```{r}
Seurat_object = CreateSeuratObject(counts = data, project = "NelsonEtAl2020")
rm(data)

Seurat_object@meta.data <- Seurat_object@meta.data %>% rownames_to_column("Assay") %>% left_join(design, by="Assay") %>% column_to_rownames("Assay")

saveRDS(Seurat_object, file = paste0(source.output.dir, "/", Seurat_object_name, ".rds"))

rm(design, study_metadata, ID_symbol_map)
```

```{r}
Seurat_metadata <- Seurat_object@meta.data
```

# Inspecting dataset, QC plots
```{r}
# Check whether mitochondrial genes are present in the samples
str_detect(rownames(Seurat_object), "^MT-") %>% sum()
```

```{r}
# Compute the % of mitochondrial counts
Seurat_object[["percent.mt"]] <- PercentageFeatureSet(Seurat_object, pattern = "^MT-") %>% mutate_all(~replace(., is.na(.), 0))

```

```{r}
Idents(Seurat_object) <- "Sample.Characteristic.individual."

# QC plots based on number of counts, features and % of mitochondrial genes
VlnPlot(Seurat_object, features = c("nFeature_RNA"))
VlnPlot(Seurat_object, features = c("nCount_RNA"))
VlnPlot(Seurat_object, features = c("percent.mt"))

FeatureScatter(Seurat_object, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(Seurat_object, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

Seurat_object@meta.data %>% select(nCount_RNA, nFeature_RNA, percent.mt) %>% summary()
```
```{r}
Seurat_object
Seurat_object_f <- subset(Seurat_object, subset = nFeature_RNA > 200 & percent.mt < 20 & nCount_RNA > 1000  & nCount_RNA < 40000)
Seurat_object_f

# QC plots based on number of counts, features and % of mitochondrial genes
VlnPlot(Seurat_object_f, features = c("nFeature_RNA"))
VlnPlot(Seurat_object_f, features = c("nCount_RNA"))
VlnPlot(Seurat_object_f, features = c("percent.mt"))

Seurat_object_f@meta.data %>% select(nCount_RNA, nFeature_RNA, percent.mt) %>% summary()
```


```{r}
Seurat_object <- Seurat_object_f
rm(Seurat_object_f)
```

# Edit metadata
```{r}
colnames(Seurat_object[[]])
table(Seurat_object$orig.ident)
table(Seurat_object$Sample.Characteristic.individual.)
table(Seurat_object$Sample.Characteristic.sex.)
table(Seurat_object$Sample.Characteristic.disease.)
table(Seurat_object$Sample.Characteristic.tumor.grading.)
table(Seurat_object$Sample.Characteristic.metastatic.site.)
table(Seurat_object$Sample.Characteristic.cell.type.)
```
```{r}
Seurat_object@meta.data <- Seurat_object[[]] %>% mutate(Project = orig.ident, Sample = Sample.Characteristic.individual.)
```

```{r}
saveRDS(Seurat_object, file = paste0(source.output.dir, "/Seurat_object.rds"))

# Seurat_object <- readRDS(file = paste0(source.output.dir, "/Seurat_object.rds"))

```

# Normalize and scale
```{r Normalize and scale}
median_library_size <- Seurat_object@assays$RNA %>% colSums() %>% median()

# Normalize dataset
Seurat_object_n <- NormalizeData(Seurat_object, normalization.method = "LogNormalize", scale.factor = median_library_size)

# find highly variable features with a variance stabilizing transform
Seurat_object_n <- FindVariableFeatures(Seurat_object_n, selection.method = "vst", nfeatures = 2000)

VariableFeaturePlot(Seurat_object_n) %>% LabelPoints(points = head(VariableFeatures(Seurat_object_n), 30), repel = TRUE)

# Scale dataset
Seurat_object_n <- ScaleData(Seurat_object_n, features = rownames(Seurat_object_n))

rm(Seurat_object)
```
# Perform PCA
```{r Perform PCA}
# Perform PCA
n_pcs = 200

Seurat_object_n <- RunPCA(Seurat_object_n, features = VariableFeatures(object = Seurat_object_n), npcs = n_pcs)

# Plot PCA results
Idents(Seurat_object_n) <- "Sample"
VizDimLoadings(Seurat_object_n, dims = 1:2, reduction = "pca")
DimPlot(Seurat_object_n, reduction = "pca")
ElbowPlot(Seurat_object_n, ndims = n_pcs)

DimHeatmap(Seurat_object_n, dims = 1:10, cells = 500, balanced = TRUE)

# Check the total variance explained
# On Seurat 3:
pca <- Seurat_object_n[["pca"]]
# Get the total variance of the scaled counts matrix:
mat <- Seurat::GetAssayData(Seurat_object_n, assay = "RNA", slot = "scale.data")
mat <- mat[rownames(mat) %in% VariableFeatures(Seurat_object_n), ]

total_variance <- sum(matrixStats::rowVars(mat))

eigValues = (pca@stdev)^2  ## EigenValues
varExplained = eigValues / total_variance

# varExplained
cumvar <- cumsum(varExplained) %>% as.data.frame() 
colnames(cumvar) <- c("cumvar")

cumvar %>% ggplot(aes(x = c(1:n_pcs), y = cumvar)) + geom_point() + labs(x = "PC #", y = "Cumulative Variance")

rm(mat, pca, cumvar)
```

# Cell cycle assignment
```{r Cell cycle assignment}
# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Assign and inspect cell cycle scores
Seurat_object_n <- CellCycleScoring(Seurat_object_n, s.features = s.genes, g2m.features = g2m.genes)

Seurat_object_n[[]] %>% select(S.Score, G2M.Score, Phase) %>% mutate_if(is.character, as.factor) %>% summary()
```

# Run UMAP and TSNE
```{r Run UMAP and TSNE}
# Run UMAP and project in 3D dimensions
Seurat_object_n <- RunUMAP(Seurat_object_n, dims = 1:n_pcs, n.components = 3L)

# Store this UMAP as a custom reduction
Seurat_object_n[["umap3d"]] <- CreateDimReducObject(embeddings = Seurat_object_n@reductions$umap@cell.embeddings, key = "umap3d_", assay = DefaultAssay(Seurat_object_n))

# Dimensionality reduction with UMAP and tSNE on PCs
Seurat_object_n <- RunUMAP(Seurat_object_n, dims = 1:n_pcs)

Seurat_object_n <- RunTSNE(Seurat_object_n, dims.use = 1:n_pcs, reduction.use = "pca", perplexity = 30)
```

# Clustering cells
```{r Clustering cells}
Seurat_object_n <- FindNeighbors(Seurat_object_n, features = VariableFeatures(object = Seurat_object_n))
Seurat_object_n <- FindClusters(Seurat_object_n, resolution = 0.25)
```

# Visualization
```{r Visualization}
plot_features <- c("Sample", "orig.ident", "Phase", "annotation", "Cell_type", "Cell_subtype", "Simplified_celltype", "Cell_class", "seurat_clusters")

plot_features <- plot_features[plot_features %in% colnames(Seurat_object_n[[]])]

for (i in plot_features){
  fig <- DimPlot(Seurat_object_n, reduction = "umap", group.by = i)
  print(fig)
}

for (i in plot_features){
  fig <- DimPlot(Seurat_object_n, reduction = "tsne", group.by = i)
  print(fig)
}

rm(fig)
```


# Annotation for inferCNV
```{r}
annotation_metadata <- Seurat_object_n@meta.data %>% mutate(
  Cell_type = ifelse(seurat_clusters %in% c(0,1,2,3), "tumor", "stromal"),
  annotation = ifelse(Cell_type == "tumor", paste(Cell_type, Sample
, sep = "_"), Cell_type))

Seurat_object_n@meta.data <- Seurat_object_n@meta.data %>% mutate(
  Cell_type = ifelse(seurat_clusters %in% c(0,1,2,3), "tumor", "stromal"),
  annotation = ifelse(Cell_type == "tumor", paste(Cell_type, Sample.Characteristic.individual.
, sep = "_"), Cell_type))

table(Seurat_object_n@meta.data$annotation)


# Write annotations file
Seurat_object_n@meta.data %>% select(annotation) %>% rownames_to_column("sample_id") %>% write.table(file = "cellAnnotations.txt", sep = "\t", row.names = FALSE, col.names = FALSE)

ref_group_names <- Seurat_object_n@meta.data %>% select(annotation) %>% unique() %>% filter(!str_detect(annotation, "tumor"))

ref_group_names %>% write.csv2(file = "ref_group_names.csv", row.names = FALSE)
```

# Plotting signatures
```{r}
DimPlot(Seurat_object_n, reduction = "umap", group.by = "annotation")
```

# Plotting signatures
```{r  Plotting signatures}
# Plotting EMT signature # Supp table 5f
EMT_signature <- c("EZH", "JUN", "VIM", "STEAP1", "SOX4", "MMP14", "SHH", "TIMP1", "ZEB1", "ZEB2", "SNAI2")

EMT_signature <- EMT_signature[EMT_signature %in% rownames(Seurat_object_n@assays$RNA)]

Idents(Seurat_object_n) <- "annotation"
for(i in 1:length(EMT_signature)){
p <- VlnPlot(Seurat_object_n, features = EMT_signature[i])
print(p)
rm(p)
}
```

```{r}
# Plotting Genes enriched in the "M" Population from Bakhoum et al Extended figure 6b 
M_markers = c("ITGB5", "ITGB1", "ITGA5", "ITGA10", "IGFBP4", "FN1", "DSC2", "CXCL1","CTNNB1", "BMP4", "BCL2L1", "VIM", "TIMP1", "TGFBR2", "TGFBI", "TGFB1", "PPARG", "NUPR1", "MSN", "LMCD1", "LEF1", "JAG1")

M_markers <- M_markers[M_markers %in% rownames(Seurat_object_n@assays$RNA)]

Idents(Seurat_object_n) <- "annotation"
for(i in 1:length(M_markers)){
p <- VlnPlot(Seurat_object_n, features = M_markers[i])
print(p)
rm(p)
}
```

```{r}
# Plot CIN signature from Bakhoum et al 2018 (Supp. table 5), comparing CIN high and CIN low in bulk data
CIN_signature <- c('PELI2','BMP2','SHH','TNS4','RAB3B','ROBO1','ARHGAP28','CHN2','CST1','F13A1','CPVL','SEMA6D','NHSL2','GTF2IP7','DPYSL3','PCDH7','KHDRBS3','TRAC','TMEM156', 'CST4','CD24','FGF5','NTN4')

CIN_signature <- CIN_signature[CIN_signature %in% rownames(Seurat_object_n@assays$RNA)]

Idents(Seurat_object_n) <- "annotation"
for(i in 1:length(CIN_signature)){
p <- VlnPlot(Seurat_object_n, features = CIN_signature[i])
print(p)
rm(p)
}

```

```{r}
# Creative Biolabs biomarkers 
#Bhattacharya M, Barlow JJ. Tumor markers for ovarian cancer. Int Adv Surg Oncol. 1979;2:155-76. PMID: 95453.
Ovarian_cancer_markers <- c("L1CAM", "CD24", "ADAM10", "BSG", "CLDN1", "CLDN3", "CLDN6", "CLDN10", "CLDN12", "CLDN17", "CLDN19", "MUC16", "BRCA1", "BRCA2", "APC", "FOLR1", "COL18A1", "CCND1", "MSLN", "WFDC2", "VEGFC", "AFP", "CGB3", "PZP")
#	miR-21 miR-141	miR-200a	miR-200b	miR-200c	miR-203	miR-205  miR-241	hMLH1	hMSH2 p14ARF	p16INK4a	DAPKinase	 FLJ12988	miR-214	miR-199a	miR-140   miR-125b1	miR-92	miR-93	miR-126	miR-29a	miR-155   miR-127	miR-99b"

Ovarian_cancer_markers <- Ovarian_cancer_markers[Ovarian_cancer_markers %in% rownames(Seurat_object_n@assays$RNA)]

Idents(Seurat_object_n) <- "seurat_clusters"
for(i in 1:length(Ovarian_cancer_markers)){
p <- VlnPlot(Seurat_object_n, features = Ovarian_cancer_markers[i])
print(p)
rm(p)
}
```


```{r}
for(i in 1:length(Ovarian_cancer_markers)){
p <- FeaturePlot(Seurat_object_n, features = Ovarian_cancer_markers[i])
print(p)
rm(p)
}
```

```{r}
#Novus biological, Muhl et al 2020., Agorku et aç. 2019
Fibroblast_signature <- c("CD39", "PLAUR", "CD44", "ITGA2", "FAS", "LY6H", "ACTA2", "FAP", "S100A4", "SOS2", "PDGFRA", "PDGFRB", "CEACAM1", "CSPG4", "P4HB")

Fibroblast_signature <- Fibroblast_signature[Fibroblast_signature %in% rownames(Seurat_object_n@assays$RNA)]

Idents(Seurat_object_n) <- "seurat_clusters"
for(i in 1:length(Fibroblast_signature)){
p <- VlnPlot(Seurat_object_n, features = Fibroblast_signature[i])
print(p)
rm(p)
}
```


```{r}
for(i in 1:length(Fibroblast_signature)){
p <- FeaturePlot(Seurat_object_n, features = Fibroblast_signature[i])
print(p)
rm(p)
}
```

```{r}
Epithelial_signature <- c("   Claudin-1 Claudin-3 Claudin-4 Claudin-6 Claudin-10a Claudin-10b Claudin-12 Claudin-17 Claudin-19 Cornulin Cytokeratin 8 Cytokeratin 14
Cytokeratin 18 Cytokeratin 19 FoxJ1 FoxN1 KLF4 KLF5 KLF10 LRRC1/LANO TCF7L1/TCF3

A33 ACE/CD143 ALCAM/CD166 Aminopeptidase B/RNPEP Aminopeptidase Inhibitors
Aminopeptidase N/CD13 Amnionless B7-H2 B7-H3 CA125/MUC16 CA15-3/MUC-1 E-Cadherin
CD1a CD1d CD1d1 CD46 CD74 CEACAM-1/CD66a CEACAM-3/CD66d CEACAM-4 CEACAM-5/CD66e
CEACAM-6/CD66c CEACAM-7 Collagen I CTRP5/C1qTNF5 Cubilin DDR1 DDR1/DDR2 beta-Defensin 2
beta-Defensin 3 alpha-Defensin 1 alpha-Defensin 5 Endorepellin/Perlecan EpCAM/TROP1
Fas Ligand/TNFSF6 Gastrokine 1 HIN-1/SCGB3A1 Hyaluronan IGSF4C/SynCAM4 Integrin alpha 4/CD49d Integrin alpha 4 beta 1 Integrin alpha 4 beta 7/LPAM-1 JAM-A
JAM-B/VE-JAM JAM-C L1CAM Laminin-1 MFG-E8 MSPR/Ron MUC-1 MUC-19 MUC-4 Nectin-1 Nectin-2/CD112 Nectin-3 Nectin-4 Nidogen-1/Entactin Occludin  PD-L1/B7-H1 PLET-1 PlGF
Prostasin/Prss8 SLURP2 TfR (Transferrin R) UGRP1/SCGB3A2 ")
```


# Export
```{r Export}
if(export){
saveRDS(Seurat_object_n, file = paste0(source.output.dir, "/Seurat_object_n.rds"))
}

# Seurat_object_n <- readRDS(file = paste0(source.output.dir, "/Seurat_object_n.rds"))
```
# sessionInfo - Package Versions
```{r}
sessionInfo()
# packageVersion("DESeq2")
# packinfo <- installed.packages(fields = c("Package", "Version"))
# packinfo[,c("Package", "Version")]
# packinfo["graphics",c("Package", "Version")]

```

